The Effect of Oil Price on Field Production: Evidence from the Norwegian Continental Shelf

### 2015-04-06 oil_prices_rr.tex
added pdf of appendix

Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.  The planning, build-out and depletion phase of a field are conceptually unique and how oil companies operating off-shore respond to oil prices can be expected to vary substantially.

@@

However even with the question of lagged production and investment, some ambiguity exists.  \citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices off-shore producers will invest more in risky wild-cat drilling in search of new fields, but concentrate investments in lower-risk ventures, like expanding production in existing fields, when prices are low.  If this effect were to dominate, then it may even be plausible that production in existing fields reacts \emph{negatively} to increases in price.

@@

In this article, I use historical production data from all 77 currently or formerly oil-producing fields on the Norwegian continental shelf in order to estimate the effect that prices have on oil production.

@@

\FloatBarrier
\section{The effect of oil price on production: theory, simulations and empirics}

@@

	Log(production_{i,t}) & = f(production\_time_{i,t}) + \beta_1 in\_place\_oil_{i,t} \\
	 \quad & + \beta_2 oil\_price\_lags_t + \beta_3 cost\_index_{t} +  
	 f(year_t) + \epsilon_{i,t}
\label{gam_price_eqn}
\end{split}
\end{equation}

@@

||\boldsymbol{y} - \boldsymbol{X\beta} ||^2 + \lambda \int_{0}^{1} [f´´(x)]^2dx

@@

One objection to the simple model is that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic differences between the fields. To address this I can add field random effects that takes into account uncertainty based on random variation between the fields in the estimation. The smoothed function can then be written as $f(production\_time_{i,t})*Z$.  Where $Z~N(0, \sigma)$ and $\sigma$ is the variance term to be estimated.

@@

Table \ref{GAM_model_table} shows the parametric coefficient terms and other summary statistics of the regression results.  The first column shows results for the simple model with a single-dimensional smoothed function of the production profile.  The second column shows results with field-level random effects, while the third columns shows results when a two-dimensional smoothed function is used to control for production profiles across field size. The latter two models are my preferred models, providing a substantially better fit, as measured by both deviance explained and R^2. Monte Carlo Experiments, which I describe further below, also indicates that the latter two models provide consistently less-biased parametric estimates of the price terms.

@@

I(in\_place\_oil\_mill\_sm3/100)             & $0.337^{***}$  & $-0.014$       &                  \\

@@

I(price/10)                                  & $-0.006$       & $0.014$        & $0.007$          \\

@@

I(price\_l1/10)                              & $0.004$        & $0.021^{*}$    & $0.015^{*}$      \\

@@

I(price\_l2/10)                              & $0.037^{***}$  & $0.047^{***}$  & $0.048^{***}$    \\
I(price\_l3/10)                              & $0.030^{***}$  & $0.025^{**}$   & $0.033^{***}$    \\

@@

I(price\_l4/10)                              & $0.003$        & $0.043^{***}$  & $0.028^{***}$    \\

@@

I(price\_l5/10)                              & $-0.020^{**}$  & $0.012$        & $-0.001$         \\

@@

I(price\_l6/10)                              & $-0.010$       & $0.008$        & $-0.001$         \\

@@

I(price\_l7/10)                              & $-0.008$       & $0.007$        & $0.004$          \\

@@

I(price\_l8/10)                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   \\

@@

EDF: s(prod\_year,name)                      &                & $33.002^{***}$ &                  \\

@@

For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}. In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant effects estimated on the 1st to 4th lags, with a peak at the 2nd lag.  F-tests of restricted models without the concurrent and 5-8th lag confirm that these terms do not significantly improve the fit of the models.

@@

For computationally intensive Generalized Additive Models, few if any analytic results are available to confirm that the estimates are unbiased.  Instead, I rely on Monte Carlo experiments to test for biased results.  Here, I create artificial field data using a cumulative logistic function with a linear price term and random variation. I then estimate the effect of price repeatedly, generating a new random component before each new estimation. Details of the Monte Carlo experiment can be found in the online appendix \url{https://github.com/jmaurit/oil_prices/blob/master/mc_sim.html}.

@@

\section{Conclusion}

@@

The modest estimated effect of prices on production adds weight to the argument of \citet{hamilton_oil_2012} that most of the increased supply of oil that comes from higher prices is from expanding the geographic and technological boundaries of oil production.  For example exploration of deep-water oil deposits off the coast of Brazil and extraction of oil sands in western Canada.

\section{Software}
I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.  I use the R package ggplot2 for plotting \citep{wickham_ggplot2:_2009}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.

@@

\FloatBarrier
\end{document}

>>>>

% Moreover, studies within the economics literature fail to take into account how oil prices have varying effects on the different stages of field development and production.  The planning, build-out and depletion phase of a field are conceptually unique and how oil companies operating off-shore respond to oil prices can be expected to vary substantially.

@@

However even with the question of lagged production and investment, some ambiguity exists.  \citet{mohn_efforts_2008} suggests and finds evidence for the idea that in periods of high oil prices off-shore producers will invest more in risky wild-cat drilling in search of new fields, but concentrate investments in lower-risk ventures, like expanding production in existing fields, when prices are low.  If this effect were to dominate, then it may even be plausible that a fall in prices leads to an \emph{increase} in production.

@@

In this article, I use historical production data from all 77 currently or formerly oil-producing fields on the Norwegian Continental Shelf in order to estimate the effect that prices have on oil production.

@@

\section{The effect of oil price on production: theory, simulations and empirics}

@@

	Log(production_{i,t}) & = f(production\_time_{i,t}) + f(year_t) + \beta_1 in\_place\_oil_{i,t}\\
	 \quad & + \beta_3 cost\_index_{t} + \mathbf{\beta_2 oil\_price\_lags_t} +  
	  \epsilon_{i,t}
\label{gam_price_eqn}
\end{split}
\end{equation}

@@

\| \mathbf{y} - \mathbf{X\beta} \| ^2 + \lambda \int_{0}^{1} [f^{\prime \prime}(x)]^2dx

@@

One objection to the simple model is that a fixed effect controlling for the general production profile of oil fields does not sufficiently address idiosyncratic differences between the fields. To address this I can add field random effects that takes into account uncertainty based on random variation between the fields in the estimation. The smoothed function can then be written as $f(production\_time_{i,t}*Z)$.  Where $Z\~N(0, \sigma)$ and $\sigma$ is the variance term to be estimated.

@@

Table \ref{GAM_model_table} shows the parametric coefficient terms and other summary statistics of the regression results.  The first column shows results for the simple model with a single-dimensional smoothed function of the production profile.  The second column shows results with field-level random effects, while the third columns shows results when a two-dimensional smoothed function is used to control for production profiles across field size. The latter two models are my preferred models, providing a substantially better fit, as measured by both deviance explained and $R^2$. Monte Carlo Experiments, which I describe further below, also indicates that the latter two models provide consistently less-biased parametric estimates of the price terms.

@@

in\_place\_oil\_mill\_100sm3             & $0.337^{***}$  & $-0.014$       &                  \\

@@

price/10                                  & $-0.006$       & $0.014$        & $0.007$          \\

@@

price\_l1/10                              & $0.004$        & $0.021^{*}$    & $0.015^{*}$      \\

@@

price\_l2/10                              & $0.037^{***}$  & $0.047^{***}$  & $0.048^{***}$    \\
                                             & $(0.009)$      & $(0.010$      & $(0.006)$        \\
price\_l3/10                              & $0.030^{***}$  & $0.025^{**}$   & $0.033^{***}$    \\

@@

price\_l4/10                              & $0.003$        & $0.043^{***}$  & $0.028^{***}$    \\

@@

price\_l5/10                              & $-0.020^{**}$  & $0.012$        & $-0.001$         \\

@@

price\_l6/10                              & $-0.010$       & $0.008$        & $-0.001$         \\

@@

price\_l7/10                              & $-0.008$       & $0.007$        & $0.004$          \\

@@

price\_l8/10                              & $-0.044^{***}$ & $-0.011$       & $-0.027^{***}$   \\

@@

EDF: s(prod\_year,re\_field)                      &                & $33.002^{***}$ &                  \\

@@

For convenience the estimated coefficients on the price terms from the two preferred models are shown in figure \ref{price_coefficients}. In general the results from the two models agree with each other, indicating little to no effect in the concurrent price term, but with significant coefficients estimated on the 1st to 4th lags, with a peak at the 2nd lag.  F-tests with comparisons to restricted models without the concurrent and 5-8th lag confirm that these terms do not significantly improve the fit of the models.

@@

For computationally intensive Generalized Additive Models, few if any analytic results are available to confirm that the estimates are unbiased.  Instead, I rely on Monte Carlo experiments to test for biasedness.  Here, I create artificial field data using a cumulative logistic function with a linear price term and random variation. I then estimate the effect of price repeatedly, generating the random component before each new estimation. Details of the Monte Carlo experiment can be found in the on-line appendix \url{https://github.com/jmaurit/oil_prices/blob/master/mc_sim.html}.

@@

\section{Conclusion and Discussion}

@@

One important caveat is that I do not claim to capture all possible mechanisms for price to influence production.  Mechanisms that are likely gradual and smooth in nature will be absorbed by the smooth functions in the regression. In particular, price-induced technological change will likely be a smooth process with significant lag times. Referring back to figure \ref{smooths}Perhaps some evidence of this can be seen in the estimated smooth function of the calendar year variable, which shows a slight (and statistically insignificant) upwards slope after approximately 2005, corresponding to a preceding period of rising oil prices.

The findings in this paper can provide some guidance on the effects of the recent steep fall in the oil price on Norwegian production, as well as production from other off-shore production areas. The fall in total oil production that has continued since the peak in production in 2000 has recently been arrested as several new finds began production after years of increased searching and investment.  As mentioned, a further significant boost is expected in 2019 when production from the large Sverdrup field begins production. The results of this paper indicate that the recent fall in prices is unlikely to have major ramifications for production in existing fields.  Anecdotal evidence however suggests that searching and exploration are being curtailed. Anecdotal evidence also suggests that lower oil prices may be leading to earlier shut-down decisions of depleting fields - a hypothesis that could be explored with future research.

Finally, relating back to the literature on total oil supply, the modest estimated effect of prices on production adds weight to the argument of \citet{hamilton_oil_2012} that most of the increased supply of oil that comes from higher prices is from expanding the geographic and technological boundaries of oil production.  For example exploration of deep-water oil deposits off the coast of Brazil and extraction of oil sands in western Canada.

\FloatBarrier
\section{Software}
I use the R statistical programming package for all the analysis in this article \citep{r_core_team_r:_2013}.  I use the R package ggplot2 for plotting \citep{wickham_ggplot2:_2009}, plyr for data manipulation and cleaning \citep{wickham_split-apply-combine_2011}, texreg for table formatting \citep{leifeld_texreg:_2013} and mgcv for implementation of the Generalized Additive Models \citep{wood_fast_2011}.

@@

\end{document}
